RTS AI: Problems and Techniques

### 2014-12-29 ECGG_RTSProblems/ECGG_RTSProblems.tex
done a complete pass addressing Florian s comments and adding some newer references

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Title
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{RTS AI Problems and Techniques
%\thanks{Supported by ...}
}

@@

\section{Introduction}\label{sec:intro}
The  field of  Real-Time Strategy  (RTS) game  Artificial Intelligence
(AI) has advanced significantly since Michael Buro's call for research
in this area \cite{Buro03rts}. Specially, competitions like the ``ORTS
RTS  Game AI  Competition''  (held  from 2006  to  2009), the  ``AIIDE
StarCraft AI Competition'' (held since  2010), and the ``CIG StarCraft
RTS AI Competition'' (held since  2011) have motivated the exploration
of many  AI approaches  in the  context of  RTS AI.  We will  list and
classify these approaches, explain  their strengths and weaknesses and
conclude on what is left to achieve human-level RTS AI.

Complex  dynamic  environments,  where neither  perfect  nor  complete
information  about the  current state  or  about the  dynamics of  the
environment are available, pose  significant challenges for artificial
intelligence. Road traffic, finance, or weather forecasts are examples
of such large, complex, real-life  dynamic environments. RTS games can
be seen  as a simplification  of one such real-life  environment, with
simpler dynamics in a finite and smaller world, although still complex
enough to  study some  of the key  interesting problems  like decision
making under  uncertainty or  real-time adversarial  planning. Finding
efficient techniques for tackling these problems on RTS games can thus
benefit other  AI disciplines and  application domains, and  also have
concrete and direct applications in the ever growing industry of video
games.

@@

This article aims to  provide an overview on what is  the state of the
art  in  RTS AI,  with  a  particular emphasis  on  the  work done  in
StarCraft.  It is organized as follows: 
{\color{blue} TODO (or not)}

@@

Real-Time Strategy is a sub-genre of strategy games where players need
to  build an  economy (gathering  resources and  building a  base) and
military power (training units  and researching technologies) in order
to defeat  their opponents  (destroying their army  and base).  From a
theoretical point of view, the  main differences between RTS games and
traditional board games such as Chess or Go are:

\begin{itemize}
\item  They are  {\em simultaneous  move} games,  where more  than one
  player  can issue  actions  at the  same  time. Additionally,  these
  actions are {\em durative}, i.e.  actions are not instantaneous, but
  take some amount of time to complete.
\item RTS games  are ``real-time'', which actually means  is that each
  player  has  a  very  small  amount  of  time  to  decide  the  next
  move. Compared to  Chess, where players may have  several minutes to
  decide the next action, in StarCraft, the game executes at 24 frames
  per second, which means that players  can act as fast as every 42ms,
  before the game state changes.
\item Most  RTS games are  partially observable: players can  only see
  the part of the  map that has been explored. This  is referred to as
  the {\em fog-of-war}.
\item Most RTS games are non-deterministic. Some actions have a chance
  of success.
\item And  finally, the complexity  of these  games, both in  terms of
  state space size and in terms of number of actions available at each
  decision cycle is very large. For  example, the state space of Chess
  is typically  estimated to  be around  $10^{50}$, heads  up no-limit
  Texas holdem  poker around $10^{80}$,  and Go around  $10^{170}$. In
  comparison,  the  state space  of  StarCraft  in  a typical  map  is
  estimated to be  many orders of magnitude larger than  any of those,
  as discussed in the next section.
\end{itemize}

For those reasons, standard techniques  used for playing classic board
games, such as  game tree search, cannot be directly  applied to solve
RTS games without the definition of some level of abstraction, or some
other simplification. Interestingly enough, humans  seem to be able to
deal with the  complexity of RTS games, and are  still vastly superior
to       computers      in       these       types      of       games
\cite{burochurchill2012aimagazine}.   For  those   reasons,  a   large
spectrum of techniques  have been attempted to deal  with this domain,
as we will describe below. The remainder of this section is devoted to
describe StarCraft  as a research  testbed, and on detailing  the open
challenges in RTS game AI.

\subsection{StarCraft}\label{subsec:StarCraft}

{\color{blue}  Do we  need this  part?  Should  we talk  about RTS  AI
  problems  and   techniques  in  general,  but   take  examples  from
  StarCraft only, arguing this game is mainstream?}

{\em StarCraft: Brood  War} is an immensely popular  RTS game released
in   1998  by   Blizzard  Entertainment.   StarCraft  is   set  in   a
science-fiction based universe where the player must choose one of the
three  races: Terran,  Protoss or  Zerg.  One of  the most  remarkable
aspects  of StarCraft  is  that  the three  races  are extremely  well
balanced.

% \begin{itemize}
% \item Terrans provide  units that are versatile and  flexible giving a
%   balanced option between Protoss and Zergs.
% \item  Protoss   units  have   lengthy  and   expensive  manufacturing
%   processes, but they are strong and resistant.  These conditions make
%   players follow a strategy of quality over quantity.
% \item Zergs, the insectoid race, units are cheap and weak. They can be
%   produced fast, encouraging players to overwhelm their opponents with
%   sheer numbers.
% \end{itemize}

% What about copyright problems with this screenshot?
% Figure~\ref{fig:StarCraft} shows  a screenshot of StarCraft  showing a
% player playing the Terran race.

{\color{blue}  Thus,  merge the  following  paragraph  with the  first
  paragraph of the RTS section?}
In order to win a StarCraft  game, players must first gather resources
(minerals  and Vespene  gas). As  resources become  available, players
need to allocate them for creating more buildings (which reinforce the
economy, and allow players to  create units or unlock stronger units),
research  new technologies  (in order  to  use new  unit abilities  or
improve the units)  and train attack units. Units  must be distributed
to  accomplish different  tasks  such as  reconnaissance, defense  and
attack.  While  performing all  of those tasks,  players also  need to
strategically understand the geometry of the  map at hand, in order to
decide where to place new buildings  (concentrate in a single area, or
expand  to  different  areas)  or where  to  set  defensive  outposts.
Finally, when  offensive units of  two players meet, each  player must
quickly maneuver each  of the units in order to  fight a battle, which
requires quick and reactive control of each of the units.

@@

From a theoretical point of view,  the state space of a StarCraft game
for a  given map is  enormous.  For  example, consider a  typical $128
\times 128$ map. At any given moment  there might be between 50 to 400
units in  the map, each of  which might have a  complex internal state
(remaining energy  and hit-points, action being  executed, etc.). This
quickly leads to an immense number  of possible states (way beyond the
size  of smaller  games,  such  as Chess  or  Go).  For example,  just
considering the location of each  unit (with $128 \times 128$ possible
positions per  unit), and  400 units,  gives us  an initial  number of
$16384^{400} \approx 10^{1685}$. If we add the other factors playing a
role in the game, we obtain even larger numbers.

@@

The reader can find a good  presentation of task decomposition for AIs
playing  RTS   in~\cite{weber2011acs}.  Although  the   previous  task
decomposition  is  common, a  significant  challenge  is on  designing
architectures so that  the individual AI techniques  that address each
of  those  tasks  can   communicate  and  effectively  work  together,
resolving conflicts, prioritizing resources between them, etc. Section
\ref{sec:bot}  provides an  overview of  the task  decompositions that
state-of-the-art bots use.  Moreover, we would like to  point out that
the task decomposition  above is not the only  possible approach. Some
systems, such  as IMAI  \cite{miles2007co}, divide gameplay  into much
smaller  tasks, which  are then  assigned resources  depending on  the
expected benefits of achieving each task.

@@

% HARD-CODED:
\subsubsection{Hard-coded  approaches}
They have  been extensively  used in commercial  RTS games.   The most
common      ones     use      finite     state      machines     (FSM)
\cite{FSM_AIGameProgWisdom2003}  in   order  to  let  the   AI  author
hard-code the strategy  that the AI will employ. The  idea behind FSMs
is to decompose the AI behavior into easily manageable states, such as
``attacking'', ``gathering resources''  or ``repairing'' and establish
the  conditions that  trigger  transitions  between them.   Commercial
approaches also include Hierarchical FSMs,  in which FSMs are composed
hierarchically.    These  hard-coded   approaches   have  achieved   a
significant  amount of  success,  and,  as we  will  discuss later  in
Section~\ref{sec:bot}, have  also been  used in  many academic  RTS AI
research systems.   However, these  hard-coded approaches  struggle to
encode  dynamic, adaptive  behaviors,  and are  easily exploitable  by
adaptive opponents.

@@

% WALLING
\subsubsection{Walling}
Walling is the act of  intentionally placing buildings at the entrance
of your  base to block  the path and  to prevent the  opponent's units
from getting inside. This technique is used by human StarCraft players
to   survive  early   aggression   and  earn   time   to  train   more
units.  \v{C}ertick\'{y} solved  this constraint  satisfaction problem
using Answer Set Programming (ASP) \cite{certicky2013wallin}.

@@

% GAME TREE SEARCH:
\subsubsection{Game  tree search}
These  techniques  have  also  been  explored  for  tactical  decision
making. Churchill and Buro \cite{churchill2012fast} presented the ABCD
algorithm  (Alpha-Beta  Considering  Durations), a  game  tree  search
algorithm   for   tactical   battles   in   RTS   games.    Chung   et
al. \cite{Chung05} applied Monte-Carlo  planning to a capture-the-flag
version  of Open  RTS.   Balla  and Fern  \cite{UCT}  applied the  UCT
algorithm (a  Monte Carlo Tree  Search algorithm) to  tactical assault
planning in Wargus. To make game tree search applicable at this level,
abstract game  state representations are  used in order to  reduce the
complexity. Also,  abstractions, or  simplifications about the  set of
possible actions to execute in a given game state need to be used.

@@

Holistic approaches  to address  RTS AI attempt  to address  the whole
problem using a  single unified method. To the best  of our knowledge,
with a few  exceptions, such as the  Darmok system \cite{OntanonMSR10}
(which uses  a combination of  case-based reasoning and  learning from
demonstration) or ALisp \cite{Marthi05}, there  has not been much work
in this  direction.  The  main reason  is that  the complexity  of RTS
games is  too large,  and approaches that  decompose the  problem into
smaller,    separate,   problems,    achieve    better   results    in
practice. However,  holistic approaches, based, for  example, on Monte
Carlo  Tree  Search,  have  only  been  explored  in  the  context  of
smaller-scale RTS games~\cite{ontanon2013naive}. Techniques that scale
up to large RTS games as StarCraft are still not available.

@@

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Title
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{RTS AI: Problems and Techniques
%\thanks{Supported by ...}
}

@@

\section{Introduction}\label{sec:intro}
The  field of  Real-Time Strategy  (RTS) game  Artificial Intelligence
(AI) has advanced significantly since Michael Buro's call for research
in this area \cite{Buro03rts}. Specially, competitions like the ``ORTS
RTS  Game AI  Competition''  (held  from 2006  to  2009), the  ``AIIDE
StarCraft AI Competition'' (held since  2010), and the ``CIG StarCraft
RTS AI Competition'' (held since  2011) have motivated the exploration
of many AI approaches  in the  context of  RTS AI. This paper, based on the work work presented in \cite{ontanon2013survey}, lists the open problems in RTS game AI, and the approaches that have been proposed to address these problems.

Complex  dynamic  environments,  where neither  perfect  nor  complete
information  about the  current state  or  about the  dynamics of  the
environment are available, pose  significant challenges for artificial
intelligence. Road traffic, finance, or weather forecasts are examples
of such large, complex, real-life  dynamic environments. RTS games can
be seen  as a simplification  of one such real-life  environments, with
simpler dynamics in a finite and smaller world, although still complex
enough to  study some  of the key  interesting problems  like decision
making under  uncertainty or  real-time adversarial  planning. Finding
efficient techniques for tackling these problems on RTS games can thus
benefit other  AI disciplines and  application domains, and  also have
concrete and direct applications in the ever growing industry of video
games.

@@

%This article aims to  provide an overview on what is  the state of the
%art  in  RTS AI,  with  a  particular emphasis  on  the  work done  in
%StarCraft.  It is organized as follows: 
%{\color{blue} TODO (or not)}

@@

Real-Time Strategy is a sub-genre of strategy games where players need
to  build an  economy (gathering  resources and  building a  base) and
military power (training units  and researching technologies) in order
to defeat  their opponents  (destroying their army  and base).  From a
theoretical point of view, the  main differences between RTS games and
traditional board games such as Chess are:

\begin{itemize}
\item  RTS games are  {\em simultaneous  move} games,  where more  than one
  player  can issue  actions  at the  same  time. 
\item Action in RTS games are {\em durative}, i.e.  actions are not instantaneous, but
  take some amount of time to complete.
\item RTS games  are ``real-time'', which actually means  is that each
  player  has  a  very  small  amount  of  time  to  decide  the  next
  move. Compared to  Chess, where players may have  several minutes to
  decide the next action, in StarCraft, a popular RTS game, the game executes at 24 frames
  per second, which means that players  can act as fast as every 42ms,
  before the game state changes.
\item Most  RTS games are  partially observable: players can  only see
  the part of the  map that has been explored. This  is referred to as
  the {\em fog-of-war}.
\item Most RTS games are non-deterministic: some actions have a chance
  of success, and the amount of damage dealt by different units is sometimes stochastic.
\item And  finally, the complexity  of these  games, both in  terms of
  state space size and in terms of number of actions available at each
  decision cycle is very large. For  example, the state space of Chess
  is typically  estimated to  be around  $10^{50}$, heads  up no-limit
  Texas holdem  poker around $10^{80}$,  and Go around  $10^{170}$. In
  comparison,  the  state space  of  StarCraft  in  a typical  map  is
  estimated to be  many orders of magnitude larger than  any of those,
  as discussed in the next section.
\end{itemize}

For those reasons, standard techniques  used for playing classic board
games, such as  game tree search, cannot be directly  applied to solve
RTS games without the definition of some level of abstraction, or some
other simplification. Interestingly enough, humans  seem to be able to
deal with the  complexity of RTS games, and are  still vastly superior
to       computers      in       these       types      of       games
\cite{burochurchill2012aimagazine}.   For  those   reasons,  a   large
spectrum of techniques  have been attempted to deal  with this domain,
as we will describe below. 
%The remainder of this section is devoted to describe StarCraft  as a research  testbed, and on detailing  the open
% challenges in RTS game AI.

%\subsection{StarCraft}\label{subsec:StarCraft}
%{\color{blue}  Do we  need this  part?  Should  we talk  about RTS  AI
%  problems  and   techniques  in  general,  but   take  examples  from
%  StarCraft only, arguing this game is mainstream?}
In the past few years, {\em StarCraft: Brood War} (an immensely popular  
RTS game released in 1998 by Blizzard  Entertainment) has become the 
standard testbed for evaluating AI techniques in RTS games.
%{\em StarCraft: Brood  War} is an immensely popular  RTS game released
%in   1998  by   Blizzard  Entertainment.   
StarCraft  is   set  in   a
science-fiction based universe where the player must choose one of the
three  races: Terran,  Protoss or  Zerg.  %One of  the most  remarkable
%aspects  of StarCraft  is  that  the three  races  are extremely  well
%balanced.
%
% \begin{itemize}
% \item Terrans provide  units that are versatile and  flexible giving a
%   balanced option between Protoss and Zergs.
% \item  Protoss   units  have   lengthy  and   expensive  manufacturing
%   processes, but they are strong and resistant.  These conditions make
%   players follow a strategy of quality over quantity.
% \item Zergs, the insectoid race, units are cheap and weak. They can be
%   produced fast, encouraging players to overwhelm their opponents with
%   sheer numbers.
% \end{itemize}
%
% What about copyright problems with this screenshot?
% Figure~\ref{fig:StarCraft} shows  a screenshot of StarCraft  showing a
% player playing the Terran race.
%
%{\color{blue}  Thus,  merge the  following  paragraph  with the  first
%  paragraph of the RTS section?}
In order to win a StarCraft  game, players must first gather resources
(minerals  and Vespene  gas). As  resources become  available, players
need to allocate them for creating more buildings (which reinforce the
economy, and allow players to  create units or unlock stronger units),
research  new technologies  (in order  to  use new  unit abilities  or
improve the units)  and train attack units. Units  must be distributed
to  accomplish different  tasks  such as  reconnaissance, defense  and
attack.  While  performing all  of those tasks,  players also  need to
strategically understand the geometry of the  map at hand, in order to
decide where to place new buildings  (concentrate in a single area, or
expand  to  different  areas)  or where  to  set  defensive  outposts.
Finally, when  offensive units of  two players meet, each  player must
quickly maneuver each  of the units in order to  fight a battle, which
requires quick and reactive control of each of the units.

@@

From a theoretical point of view,  the state space of a StarCraft game
for a  given map is  enormous.  For  example, consider a  typical $128
\times 128$ map. At any given moment  there might be between 50 to 400
units in  the map, each of  which might have a  complex internal state
(remaining energy  and hit-points, action being  executed, etc.). This
quickly leads to an immense number  of possible states (way beyond the
size  of smaller  games,  such  as Chess  or  Go).  For example,  just
considering the location of each  unit (with $128 \times 128$ possible
positions per  unit), and  400 units,  gives us  an initial  number of
$16384^{400} \approx 10^{1685}$. If we add the other factors playing a
role in the game, we obtain even larger numbers (see \ref{ontanon2013survey} for a more in-depth description of the complexity of StarCraft).

@@

The reader can find a good  presentation of task decomposition for AIs
playing  RTS   in~\cite{weber2011acs}.  Although  the   previous  task
decomposition  is  common, a  significant  challenge  is on  designing
architectures so that  the individual AI techniques  that address each
of  those  tasks  can   communicate  and  effectively  work  together,
resolving conflicts, or prioritizing resources between them.% Section
%\ref{sec:bot}  provides an  overview of  the task  decompositions that
%state-of-the-art bots use.  
Moreover, we would like to  point out that
the task decomposition  above is not the only  possible approach. Some
systems, such  as IMAI  \cite{miles2007co}, divide gameplay  into much
smaller  tasks, which  are then  assigned resources  depending on  the
expected benefits of achieving each task.

@@

% HARD-CODED:
\subsubsection{Hard-coded  approaches}
They have  been extensively  used in commercial  RTS games.   The most
common      ones     use      finite     state      machines     (FSM)
\cite{FSM_AIGameProgWisdom2003}  in   order  to  let  the   AI  author
hard-code the strategy  that the AI will employ. The  idea behind FSMs
is to decompose the AI behavior into easily manageable states, such as
``attacking'', ``gathering resources''  or ``repairing'' and establish
the  conditions that  trigger  transitions  between them.   Commercial
approaches also include Hierarchical FSMs,  in which FSMs are composed
hierarchically.    These  hard-coded   approaches   have  achieved   a
significant  amount of  success,  and, have  also been  used in  many academic  RTS AI
research systems.   However, these  hard-coded approaches  struggle to
encode  dynamic, adaptive  behaviors,  and are  easily exploitable  by
adaptive opponents.

@@

% WALLING
\subsubsection{Walling}
Walling is the act of  intentionally placing buildings at the entrance
of your  base to block  the path and  to prevent the  opponent's units
from getting inside. This technique is used by human StarCraft players
to   survive  early   aggression   and  earn   time   to  train   more
units.  \v{C}ertick\'{y} addressed  this constraint  satisfaction problem
using Answer Set Programming (ASP) \cite{certicky2013wallin}. Richoux et al. \cite{richoux2014walling} presented an alternative approach based on constraint programming and local search, designed to be run-time.

@@

% GAME TREE SEARCH:
\subsubsection{Game  tree search}
These  techniques  have  also  been  explored  for  tactical  decision
making. Churchill and Buro \cite{churchill2012fast} presented the ABCD
algorithm  (Alpha-Beta  Considering  Durations), a  game  tree  search
algorithm   for   tactical   battles   in   RTS   games.    Chung   et
al. \cite{Chung05} applied Monte-Carlo  planning to a capture-the-flag
version  of Open  RTS.   Balla  and Fern  \cite{UCT}  applied the  UCT
algorithm (a  Monte Carlo Tree  Search algorithm) to  tactical assault
planning in Wargus. To make game tree search applicable at this level,
abstract game  state representations are  used in order to  reduce the
complexity. Uriarte and Onta\~{n}\'{o}n \cite{uriarte2014high,uriarte2014game} explored different game state abstractions in the context of Monte-Carlo Tree Search for high-level tactical reasoning in StarCraft. Other algorithms, such as Greedy Portfolio Search \cite{churchill2013portfolio}, perform abstraction at the level of actions, by employing a collection of predefined ``scripts'', and using these scripts as the possible actions that the players can execute in the context of game tree search.

@@

Finally, holistic approaches  to address  RTS AI attempt  to address  the whole
problem using a  single unified method. To the best  of our knowledge,
with a few  exceptions, such as the  Darmok system \cite{OntanonMSR10}
(which uses  a combination of  case-based reasoning and  learning from
demonstration) or ALisp \cite{Marthi05}, there  has not been much work
in this  direction.  The  main reason  is that  the complexity  of RTS
games is  too large,  and approaches that  decompose the  problem into
smaller,    separate,   problems,    achieve    better   results    in
practice. However,  holistic approaches, based, for  example, on Monte
Carlo  Tree  Search,  have  only  been  explored  in  the  context  of
smaller-scale RTS games~\cite{ontanon2013naive}. Techniques that scale
up to large RTS games as StarCraft are still not available.

@@

%\section{Conclusions}
%
%This paper has summarized the set of open problem in RTS game AI, and summarized existing work toward addressing those problems.
